# Reproducible Research: Peer Assessment 1

```{r, echo = FALSE}
createtime <- format(Sys.time(), "%a %b %d %X %Y")
```

This document was created `r createtime`.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and includes the number of steps taken in 5 minute intervals each day.

The data and Rmd markdown template were obtained by Forking and Cloning the GitHub repository https://github.com/rdpeng/RepData_PeerAssessment1, SHA-1 dc20c7c4e92aea6af318a611e2669d5b9c6cbfaf from February 11, 2014.

## Loading and preprocessing the data
```{r, opts_chunk$set(echo=TRUE)}
setwd("~/Desktop/RepData_PeerAssessment1")
files <- list.files()
files

## read and summarize the activity monitoring data
act1 <- read.csv(file = "activity.csv", header = TRUE)
str(act1)
summary(act1)
```

## What is mean total number of steps taken per day?
There is some ambiguity in the instructions as to whether day is referring to days of the week or date.  Let's assume the calendar date, because the instructions refer to weekdays later.

"For this part of the assignment, you can ignore the missing values in the dataset."

The assignment instructions state "Make a histogram of the total number of steps taken each day".  So here it is, but more interesting is the total number of steps by date as shown in the bar plot below


```{r fig.path='Figures/'}
# split on date
split1 <- split(act1$steps, act1$date)
totstepsday <- sapply(split1, sum)
hist(totstepsday, main = "Daily Total Steps")
#barplot(totstepsday, xlab = "date", ylab = "total steps", main = "Total Steps by Date")
```

Calculate and report the mean and median total number of steps taken per day

```{r}
meantotstepsday <- mean(totstepsday, na.rm = TRUE)
mediantotstepsday <- median(totstepsday, na.rm = TRUE)
```

The mean steps taken per day is `r meantotstepsday` and the the median is `r mediantotstepsday`.

## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).  The interval data should be converted from a numeric to time value.  0 is midnight, 500 is 5:00 am.

```{r fig.path='Figures/'}
library(reshape2)
melt1 <- melt(act1, id = c("date", "interval"), variable = "steps")
# calculate the mean steps for each interval
means1 <- dcast(melt1, interval ~ steps, fun = mean, na.rm = TRUE)
plot(x = means1$interval, y = means1$steps, type = "l", xlab = "interval", 
     ylab = "mean steps", main = "Mean Steps by Interval")
```


Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxsteps <- max(means1$steps)
maxinterval <- means1[means1$steps == maxsteps, ]
maxinterval
```

The maximum steps for a five minute interval is `r maxsteps` and is in interval `r maxinterval$interval`.


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

```{r}
numna <- sum(is.na(act1$steps))
```

The number of missing values in the dataset is `r numna`.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3.  Create a new dataset that is equal to the original dataset but with the missing data filled in.

Replace NA values in steps with mean values for each time interval
```{r}
act2 <- act1
missing <- is.na(act2$steps)
act2[missing,"steps"] <- means1$steps
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r fig.path='Figures/'}
split2 <- split(act2$steps, act2$date)
totstepsday2 <- sapply(split2, sum)
hist(totstepsday2, main = "Daily Total Steps")
#barplot(totstepsday2, xlab = "date", ylab = "total steps", main = "Total Steps by Date")
meantotstepsday2 <- mean(totstepsday2, na.rm = TRUE)
mediantotstepsday2 <- median(totstepsday2, na.rm = TRUE)
# make a nice table
t <- data.frame(row.names = c("Mean", "Median"))
t[1, 1] <- meantotstepsday
t[1, 2] <- meantotstepsday2
t[2, 1] <- mediantotstepsday
t[2, 2] <- mediantotstepsday2
names(t) = c("Original","Imputed" )
t
```

The missing values for steps for 8 days were replaced with the 288 mean interval values.  This has an impact of total steps for the entire dataset as can be seen in the histogram and barplots, but no difference in the mean total steps per day.  The median total steps per day is different by 1 step.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
dayofweek <- weekdays(as.Date(act2$date))
act3 <- cbind(act2, dayofweek)
act3[act3$dayofweek == "Monday", 5] <- "weekday"
act3[act3$dayofweek == "Tuesday", 5] <- "weekday"
act3[act3$dayofweek == "Wednesday", 5] <- "weekday"
act3[act3$dayofweek == "Thursday", 5] <- "weekday"
act3[act3$dayofweek == "Friday", 5] <- "weekday"
act3[act3$dayofweek == "Saturday", 5] <- "weekend"
act3[act3$dayofweek == "Sunday", 5] <- "weekend"
names(act3)[5] <- "type"
# check
table(act3$dayofweek)
table(act3$type)
```


2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

Two plots are provided below.  The first shows mean steps per interval in one panel color coded on weekend/weekday.  This plot allows easy comparison of the weekend to weekday and that the subject gets started later in the day on weekends.  The second plot is the two panel plot requested in the instructions for this assignment.


```{r fig.path='Figures/'}
melt3 <- melt(act3, id = c("date", "dayofweek", "type", "interval"), variable = "steps")
# calculate the mean steps for each interval by type
means3 <- dcast(melt3, interval + type ~ steps, fun = mean, na.rm = TRUE)
#library(ggplot2)
#ggplot(means3, aes(interval, steps, col = type)) + 
#    geom_line() +
#    theme_bw()
#ggplot(means3, aes(interval, steps)) +
#    geom_line() +
#    facet_wrap(~type, ncol = 1) +
#    theme_bw()
library(lattice)
xyplot(steps ~ interval | type, data = means3, type = "l", layout = c(1,2))
```

